Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-08-10 chapters/Paper_three.tex
All references polished, conclusion missing

\section{Introduction} Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{gross1980fluid,szeri2010fluid,DERYCK1998278,doi:10.1146/annurev.fluid.31.1.347,dasilvasobrinhoStudyDefectsUltrathin1999,singh2010inkjet,jo2009evaluation,Wijshoff2010}. 
Moreover, the continuously growing technological interest for lab-on-a-chip devices~\cite{C6LC00387G,Focke} as well as for printable electronics~\cite{Kim_2005, Luechinger_2008} and printable photovoltaics~\cite{Brabec2001,RH20}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{ButtEtAl_Langmuir2018,GuoGuo_RSCAdv2016}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{LiuChem2005, XiCSR2010}, involving smart materials such as polymer brushes~\cite{CohenStuartEtAl_NatMat2010,ayres2007stimuli,YongEtAl_Mat2018}, thermal-responsive hydrogels~\cite{ChenEtAl_SM2010}, light-responsive molecules and microstructures~\cite{ichimuraLightDrivenMotionLiquids2000,DelormeEtAl_Langmuir2005,OscuratoEtAl_AAMI2017} or processes such as electrowetting~\cite{MugeleEtAl_JPCM2005}.
Probably the simplest, yet non-trivial, modelling of a switchable substrate can be realized by a space and time dependent wettability pattern~\cite{D0SM02082F}. 
While a consistent body of theoretical/computational work has been devoted to processes on static heterogeneous substrates, though, the time dependent case is still almost unexplored, with few relevant exceptions focusing on single droplet spreading and sliding~\cite{D0SM02082F,GrawitterStark2,ThieleHartmann} or limited to an analysis of the linear regime~\cite{suman2006dynamics}.

@@

\section{Method} In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009} 

@@

by means of a recently developed method, based on a lattice Boltzmann (LB) \\scheme~\cite{PhysRevE.100.033313,zitzLatticeBoltzmannSimulations2021, Zitz2022}.
Eq.~(\ref{eq:thinfilm}) describes, in a lubrication approximation spirit, the evolution of the height field (film thickness) $h(\mathbf{x},t)$, denoting the location of the liquid/air interface. The mobility function 
$M_{\delta}(h) = \frac{2h^3 + 6\,\delta\, h^2 + 3\delta^2h}{6\mu}$
depends on the velocity boundary condition at the substrate, parameterized by an effective slip length $\delta$
(for $\delta \rightarrow 0$ it reduces to the no-slip form $h^3/(3\mu)$). Here, $\mu$ is the fluid dynamic viscosity.
The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$~\footnote{The Laplace pressure is the product of the surface tension 
and the local curvature, therefore its full expression reads $p_L = -\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$; 
since we are in the lubrication regime, though, whereby $|\nabla h| \ll 1$, the denominator can be safely neglected 
and the pressure reduces to the form involving only the Laplacian of the thickness 
field~\cite{Benet2014,alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}.}, 

@@

where $f(\xi)=\xi^{-3} - \xi^{-2}$ and $h_{\ast}$ is the height at which the disjoining pressure vanishes and sets the precursor layer thickness~\footnote{Expressions of 
the general form $\Pi(h) = \gamma(1-\cos \theta)(n-1)(m-1)/(n-m)((h_{\ast}/h)^n - (h_{\ast}/h)^m)/h_{\ast}$
are frequently found in the literature, with various combination of exponents $(n,m)$ ($n>m$), depending
on the type of interactions among fluid molecules and between fluid and substrate~\cite{schwartzSimulationDropletMotion1998,Mitlin,Teletzke}. A common choice, for instance, is the pair $(9,3)$, which 
can be derived from molecular interactions of Lennard-Jones type. Here, the use of the $(3,2)$ pair is motivated mainly by the computational advantage of keeping the characteristic dewetting time 
shorter. We have run also a single case ($\lambda=256 h_0$ and $\Gamma=1.5$) with $(9,3)$ and observed that 
the dynamics remains essentially the same, but for a shift in the time scale (see also~\cite{zitzLatticeBoltzmannSimulations2021})}.

@@

We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two 
directions~\cite{KarguptaLangmuir2003,Jalali2018,Nagayama2020,Das2020}.

@@

A possible laboratory experiment for the here mentioned approach consists of a light switchable substrate with a thin film cast on it~\cite{beckerComplexDewettingScenarios2003, IchimuraEtAl_Science2000}. 
The external stimulus, our moving pattern, can be realized, e.g., by using a digital multimirror device (DMD)~\cite{doi:10.1021/jp301092y,doi:10.1126/science.aax8760}.
We fix the velocity direction to one diagonal, namely $\mathbf{v}_{\theta} = (v_{\theta x},v_{\theta y}) = v_{\theta}(1/\sqrt{2},-1/\sqrt{2})$ (we will return later on the importance of this choice), and we set $\theta_0 = 20^{\circ}$ and $\delta\theta=10^{\circ}$~\footnote{Notice that, since a typical velocity is such that $v_{\theta} \Delta t \ll \Delta x$ (in one LB time step $\Delta t$ the wave would travel a distance much smaller than a lattice spacing $\Delta x$),
the time update needs to be interpreted in an integer part sense, that is the pattern is shifted by one $\Delta x$ every $1/v_{\theta x}$ time steps (and equivalently in the $y$-direction).}.
Lengths and time scales will be expressed, respectively, in units of the mean film height, $h_0$ (which is constant in time, due to mass conservation), and of $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}$, the inverse growth rate of the most unstable mode, whose wavenumber is $q_0$, of a spinodally dewetting film \cite{meckeThermalFluctuationsThin2005}. On a uniform substrate, with constant contact angle $\theta^{(u)}$, the wavenumber reads 
$(q^{(u)}_0)^2 = h_{\ast}^{-2}(1-\cos \theta^{(u)})f^{\prime}(h_0/h_{\ast})$ \cite{meckeThermalFluctuationsThin2005,zhangMolecularSimulationThin2019}. In our patterned case, we define 
  $q_0^2=h_{\ast}^{-2}(1-\cos\theta_0)f^{\prime}(h_0/h_{\ast})$. Correspondingly, we choose as a velocity scale $v_0 = \lambda_s/t_0$, where
$\lambda_s = 2\pi/q_0$.

@@

To give a flavour of how to compare with actual experimental systems, let us notice
first that the expression of $t_0$ depends on the disjoining pressure (through its dependence 
 $q_0 = \sqrt{\frac{1}{2\gamma} \Pi^{\prime}(h_0)}$) and contains, therefore, energetic details, namely 
 the Hamaker constant $A_H$. Consider, for instance, the problem studied in \cite{beckerComplexDewettingScenarios2003}: the dewetting of a polystyrene film of thickness $h_0 \approx 4 \, \text{nm}$ deposited on a silicon 
dioxide substrate, where $A_H  = 2.2 \times 10^{-20} \, \text{J}$ and 
$\gamma = 0.03 \, \text{N}/\text{m}$. Neglecting the short range part of the interface potential
(such that $\Pi(h)= - \frac{A_H}{6\pi h^3}$)~\cite{meckeThermalFluctuationsThin2005,beckerComplexDewettingScenarios2003}, one gets 
$q_0^2 \approx \frac{A_H}{4\pi \gamma h_0^4} \approx 2.5 \times 10^{-4} \, \text{nm}^{-2}$.
The characteristic time of the growth of the instability for these experiments, then, reads \cite{fetzerThermalNoiseInfluences2007} (the dynamic viscosity of polystyrene is 
$\mu = 1.2 \times 10^4 \, \text{Pa} \cdot \text{s}$) $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4} \approx 300 \, \text{s}$ and it can be seen that indeed film rupture occurs within time 
scales of the order of $t_0$ ($\approx 2 t_0$) \cite{beckerComplexDewettingScenarios2003}, as in our 
simulation for the largest pattern wavelength, $\lambda = L$ (see also our previous work
with homogeneous substrates~\cite{zitzLatticeBoltzmannSimulations2021}).

@@

All our simulations are run on a bi-periodic square domain of size $L \times L$ with 
$L/h_0 = 512 $. The numerical values in lattice Boltzmann units of the 
mean film thickness, dynamics viscosity and surface tension are, respectively, $h_0=1$, $\mu=1/6$
and $\gamma=0.01$.
To regularize the contact line divergence~\cite{huh1971hydrodynamic} we use a precursor layer thickness $h_{\ast}/h_0=0.07$ and a slip length $\delta/h_0 = 1$, 
a value which lies
within the weak/intermediate slip regime~\cite{peschkaSignaturesSlipDewetting2019,fetzer2007quantifying, munchLubricationModelsSmall2005} (we have tried also a value four times smaller, without observing 
any major qualitative change in the results).

@@

Various wavelengths, in the range $\lambda/h_0 \in [64, 512]$, and velocities,
$v_{\theta}/v_0 \in [0, 200]$, are considered for the wettability pattern, Eq.~(\ref{eq:sinetheta}). 
A few comments on these ranges are noteworthy at this point. The wavelengths are obviously limited from above by the size of the computational domain. On the other extreme, too short wavelengths are prevented by
the energetic cost of accomodating droplets on too
small patches, since the local contact angle 
(that increases with decreasing $\lambda$~\cite{SuppMat}) would possibly exceed the largest substrate contact angle, such that the patterning 
itself would loose meaning. 
It has been, in fact, proven, both theoretically and experimentally, that 
for the confinement to be effective, the pattern characteristic size should 
not be much smaller than the spinodal length scale on the less wettable substrate
\cite{KarguptaJCP2002,KarguptaLangmuir2000,Nisato1999,Karim1998}.

@@

The pattern speeds, on the other hand, are limited by the intrinsic LB bound of the lattice speed of sound
 \cite{succi}.\\
The Reynolds number, $Re$, is of the order of $Re \sim 10^{-2}$ on the 
static substrate and never exceeds the value $\approx 0.2$ in the time dependent case.

@@

\section{Results} We start our analysis studying, first, how the rupture times depend on the parameters characterising the wettability pattern, namely the wavelength of the contact angle variation, $\lambda$, and wave speed, $v_{\theta}$.

@@

It is conveyed that, overall, rupture occurs earlier on the static substrate, suggesting that the time variation tends to stabilize the film, in agreement with linear stability analysis results~\cite{suman2006dynamics}.

@@

Let us first notice that the film rupture consists of two processes, each with its associated time scale: the growth of unstable interface perturbation with rate $t_{\theta}^{-1}$, and the retraction of liquid from more hydrophobic regions, occurring in a characteristic time $t_R$. For longer pattern wavelengths, the dewetting instability amplifies more slowly, such that it is reasonable to take it as the process determining the rupture time, $\tau_r \sim t_{\theta}$.
In this case, from the linearized thin-film equation (in one spatial dimension, for simplicity), obtained setting $h=h_0 + \delta h$ with $\delta h \ll h_0$, 
we can easily see that the exponential growth of the height perturbation is affected by the wettability pattern (variable contact angle) in such a way that $\partial_t (\delta h) \propto (\partial_x^2 (\partial_h\Pi(h_0))) \delta h$. Therefore, since the characteristic time $t_{\theta}$ can be estimated dimensionally as $t_{\theta} \sim \delta h/\dot{(\delta h)}$, the rupture time should go as

@@

The latter is related to the time the liquid takes to flow out of regions of high contact angle, whose size is $\sim \lambda$. Hence we have 

@@

    \caption{Rupture times, $\tau_r$, as a function of the pattern wavelength $\lambda$, for $v_{\theta}=0$ (\textcolor{jlblue}{$\bullet$}) and $v_{\theta}=20 v_0$ 
    (\textcolor{jlorange}{$\star$}).
    The continuous and dashed lines indicate the linear, $\sim \lambda$, and quadratic, $\sim \lambda^2$, scaling laws, respectively.

@@

    \caption{MAIN PANEL: Time evolution of the height fluctuations, $\Delta h(t)$, during the dewetting process on the patterned substrate given by       
    Eq.~(\ref{eq:sinetheta})
    with $v_{\theta}= 0$ and $\lambda= 512 h_0$ (blue circles), $\lambda=256 h_0$ (orange squares) and $\lambda=170 h_0$ (green stars).

@@

    INSET: Number of droplets, $N(t)$, as a function of time. The three horizontal dashed lines indicate the number of minima of Eq.~(\ref{eq:sinetheta}),
      which is $2\left(\frac{L}{\lambda}\right)^2$. In the insets, we display snapshots of the stationary droplet states as grey-scale images
      of the film thickness field, $h(\mathbf{x},t)$.

@@

In the main panel of the same figure, we see that the height fluctuations $\Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}$
grow in time until film rupture and then settle to a constant value, which represents a measure of the mean droplet height $h_d$ (since
droplets are essentially monodisperse), decreasing with the pattern wavelength (as expectable, due to the decrease of the droplet volume,
$V_d = \frac{h_0 \lambda^2}{2}$).\\

@@

For $v_{\theta} = 0.2 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{D0SM02082F}.

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{doi:10.1063/1.4774084, Schaller2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Figure_5.pdf}
    \caption{MAIN PANEL: Rivulets life-times, $\tau_{\text{riv}}$, for various $\Gamma$ values.
    The dashed line is plotted as a guide to the eye to highlight the logarithmic dependence, in agreement with 
    the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, computed along the rivulet axis, for three different $\Gamma$ values.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{doi:10.1063/1.3211248, PhysRevE.77.061605}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state"~\footnote{see movie file ligament\_formation\_and\_breakup.mp4 in~\cite{SuppMat}.}.

@@

If $U_{\Theta}$ is large as compared to $v_{\theta}$, the film retraction is faster than the transport due to the contact angle field and thus droplets from. However, if $v_{\theta}$ is larger than $U_{\Theta}$, then the retracting film has too little time to form droplets and ends up in the metastable rivulet state. It appears, therefore, natural to consider the ratio of these two velocities, 

@@

The rivulet itself is, in fact, prone to dewetting, with the liquid accumulating over patches around contact angle minima. However, as the pattern moves, the instability is tamed due to configurations whereby higher contact angle regions underlie height field maxima, thus tending to revert the fluid flow.

>>>>

\section{Introduction} 
Wet surfaces and droplets are part of our every-day experience and of numerous industrial processes including coating, tribology, painting and printing, to name but a few~\cite{grossFluidFilmLubrication1980,szeriFluidFilmLubrication2010,deryckGravityInertiaEffects1998,quereFluidCoatingFiber1999,dasilvasobrinhoStudyDefectsUltrathin1999,singhInkjetPrintingProcess2010,joEvaluationJetPerformance2009,wijshoffDynamicsPiezoInkjet2010}. 
Moreover, the continuously growing technological interest for lab-on-a-chip devices~\cite{samieiReviewDigitalMicrofluidics2016,fockeLabonaFoilMicrofluidicsThin2010} as well as for printable electronics~\cite{kimHighlyConductiveInk2005, luechingerGraphenestabilizedCopperNanoparticles2008} and printable photovoltaics~\cite{brabecPlasticSolarCells2001,ronsinRoleInterplaySpinodal2020}, whose efficiency relies crucially on a precise control of material deposition upon (de-)wetting of liquid films, drew the attention to applications where the substrate is adaptive or switchable, i.e. it is not inert but responds dynamically to external stimuli or to the evolution of the coating liquid film itself~\cite{buttAdaptiveWettingAdaptation2018,guoInspiredSmartMaterials2016}. 
Several realizations of switchable and adaptive substrates have been proposed~\cite{liuControlledSwitchableSurface2005, xinReversiblySwitchableWettability2010}, involving smart materials such as polymer brushes~\cite{stuartEmergingApplicationsStimuliresponsive2010,ayresStimuliResponsivePolyelectrolytePolymer2007,yongCononsolvencyTransitionPolymer2018}, thermal-responsive hydrogels~\cite{chenThermalresponsiveHydrogelSurface2010}, light-responsive molecules and microstructures~\cite{ichimuraLightDrivenMotionLiquids2000,delormeAzobenzeneContainingMonolayerPhotoswitchable2005,oscuratoLightDrivenWettabilityTailoring2017} or processes such as electrowetting~\cite{mugeleElectrowettingConvenientWay2005}.
Probably the simplest, yet non-trivial, modelling of a switchable substrate can be realized by a space and time dependent wettability pattern~\cite{grawitterSteeringDropletsSubstrates2021}. 
While a consistent body of theoretical/computational work has been devoted to processes on static heterogeneous substrates, though, the time dependent case is still almost unexplored, with few relevant exceptions focusing on single droplet spreading and sliding~\cite{grawitterSteeringDropletsSubstrates2021,grawitterDropletsSubstratesOscillating2021,thieleGradientDynamicsModel2020} or limited to an analysis of the linear regime~\cite{sumanDynamicsThinLiquid2006}.

@@

\section{Method} 
In order to simulate the dewetting dynamics on patterned, ``switchable'', substrates, we integrate numerically the thin-film equation~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009} 

@@

by means of a recently developed method, based on a lattice Boltzmann (LB) scheme~\cite{zitzLatticeBoltzmannMethod2019,zitzLatticeBoltzmannSimulations2021, zitzSwalbeJlLattice2022}.
Eq.~(\ref{eq:thinfilm}) describes, in a lubrication approximation spirit, the evolution of the height field (film thickness) $h(\mathbf{x},t)$, denoting the location of the liquid/air interface. 
The mobility function $M_{\delta}(h) = \frac{2h^3 + 6\,\delta\, h^2 + 3\delta^2h}{6\mu}$ depends on the velocity boundary condition at the substrate, parameterized by an effective slip length $\delta$ (for $\delta \rightarrow 0$ it reduces to the no-slip form $h^3/(3\mu)$). 
Here, $\mu$ is the fluid dynamic viscosity.
The film pressure $p(\mathbf{x},t)$ consists of the sum of the Laplace and disjoining pressures, that is $p(\mathbf{x},t) = -\gamma \nabla^2 h - \Pi$~\footnote{The Laplace pressure is the product of the surface tension and the local curvature, therefore its full expression reads $p_L = -\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$; since we are in the lubrication regime, though, whereby $|\nabla h| \ll 1$, the denominator can be safely neglected and the pressure reduces to the form involving only the Laplacian of the thickness field~\cite{benetDisjoiningPressureHealing2014,pahlavanThinFilmsPartial2018}.}, 

@@

where $f(\xi)=\xi^{-3} - \xi^{-2}$ and $h_{\ast}$ is the height at which the disjoining pressure vanishes and sets the precursor layer thickness~\footnote{Expressions of the general form $\Pi(h) = \gamma(1-\cos \theta)(n-1)(m-1)/(n-m)((h_{\ast}/h)^n - (h_{\ast}/h)^m)/h_{\ast}$ are frequently found in the literature, with various combination of exponents $(n,m)$ ($n>m$), depending on the type of interactions among fluid molecules and between fluid and substrate~\cite{schwartzSimulationDropletMotion1998,mitlinDewettingSolidSurface1993,teletzkeHowLiquidsSpread1987}. 
A common choice, for instance, is the pair $(9,3)$, which can be derived from molecular interactions of Lennard-Jones type. 
Here, the use of the $(3,2)$ pair is motivated mainly by the computational advantage of keeping the characteristic dewetting time shorter. 
We have run also a single case ($\lambda=256 h_0$ and $\Gamma=1.5$) with $(9,3)$ and observed that the dynamics remains essentially the same, but for a shift in the time scale (see also~\cite{zitzLatticeBoltzmannSimulations2021})}.

@@

We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two directions~\cite{karguptaMesopatterningThinLiquid2003,jalaliFabricationCharacterizationScalable2018,nagayamaIntermediateWettingState2020,dasSurfacedirectedSpinodalDecomposition2020}.

@@

A possible laboratory experiment for the here mentioned approach consists of a light switchable substrate with a thin film cast on it~\cite{beckerComplexDewettingScenarios2003, ichimuraLightDrivenMotionLiquids2000}. 
The external stimulus, our moving pattern, can be realized, e.g., by using a digital multimirror device (DMD)~\cite{vieyrasalasActiveControlEvaporative2012,sahaScalableSubmicrometerAdditive2019}.
We fix the velocity direction to one diagonal, namely $\mathbf{v}_{\theta} = (v_{\theta x},v_{\theta y}) = v_{\theta}(1/\sqrt{2},-1/\sqrt{2})$ (we will return later on the importance of this choice), and we set $\theta_0 = 20^{\circ}$ and $\delta\theta=10^{\circ}$~\footnote{Notice that, since a typical velocity is such that $v_{\theta} \Delta t \ll \Delta x$ (in one LB time step $\Delta t$ the wave would travel a distance much smaller than a lattice spacing $\Delta x$), the time update needs to be interpreted in an integer part sense, that is the pattern is shifted by one $\Delta x$ every $1/v_{\theta x}$ time steps (and equivalently in the $y$-direction).}.
Lengths and time scales will be expressed, respectively, in units of the mean film height, $h_0$ (which is constant in time, due to mass conservation), and of $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}$, the inverse growth rate of the most unstable mode, whose wavenumber is $q_0$, of a spinodally dewetting film \cite{meckeThermalFluctuationsThin2005}. 
On a uniform substrate, with constant contact angle $\theta^{(u)}$, the wavenumber reads $(q^{(u)}_0)^2 = h_{\ast}^{-2}(1-\cos \theta^{(u)})f^{\prime}(h_0/h_{\ast})$~\cite{meckeThermalFluctuationsThin2005,zhangMolecularSimulationThin2019}. 
In our patterned case, we define $q_0^2=h_{\ast}^{-2}(1-\cos\theta_0)f^{\prime}(h_0/h_{\ast})$. 
Correspondingly, we choose as a velocity scale $v_0 = \lambda_s/t_0$, where $\lambda_s = 2\pi/q_0$.

@@

To give a flavour of how to compare with actual experimental systems, let us notice first that the expression of $t_0$ depends on the disjoining pressure (through its dependence $q_0 = \sqrt{\frac{1}{2\gamma} \Pi^{\prime}(h_0)}$) and contains, therefore, energetic details, namely the Hamaker constant $A_H$.
Consider, for instance, the problem studied in \cite{beckerComplexDewettingScenarios2003}: the dewetting of a polystyrene film of thickness $h_0 \approx 4 \, \text{nm}$ deposited on a silicon dioxide substrate, where $A_H  = 2.2 \times 10^{-20} \, \text{J}$ and $\gamma = 0.03 \, \text{N}/\text{m}$. 
Neglecting the short range part of the interface potential (such that $\Pi(h)= - \frac{A_H}{6\pi h^3}$)~\cite{meckeThermalFluctuationsThin2005,beckerComplexDewettingScenarios2003}, one gets $q_0^2 \approx \frac{A_H}{4\pi \gamma h_0^4} \approx 2.5 \times 10^{-4} \, \text{nm}^{-2}$.
The characteristic time of the growth of the instability for these experiments, then, reads \cite{fetzerThermalNoiseInfluences2007} (the dynamic viscosity of polystyrene is $\mu = 1.2 \times 10^4 \, \text{Pa} \cdot \text{s}$) $t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4} \approx 300 \, \text{s}$ and it can be seen that indeed film rupture occurs within time scales of the order of $t_0$ ($\approx 2 t_0$) \cite{beckerComplexDewettingScenarios2003}, as in our simulation for the largest pattern wavelength, $\lambda = L$ (see also our previous work with homogeneous substrates~\cite{zitzLatticeBoltzmannSimulations2021}).

@@

All our simulations are run on a bi-periodic square domain of size $L \times L$ with $L/h_0 = 512 $. The numerical values in lattice Boltzmann units of the mean film thickness, dynamics viscosity and surface tension are, respectively, $h_0=1$, $\mu=1/6$ and $\gamma=0.01$.
To regularize the contact line divergence~\cite{huhHydrodynamicModelSteady1971} we use a precursor layer thickness $h_{\ast}/h_0=0.07$ and a slip length $\delta/h_0 = 1$, a value which lies within the weak/intermediate slip regime~\cite{peschkaSignaturesSlipDewetting2019,fetzerQuantifyingHydrodynamicSlip2007, munchLubricationModelsSmall2005} (we have tried also a value four times smaller, without observing any major qualitative change in the results).

@@

Various wavelengths, in the range $\lambda/h_0 \in [64, 512]$, and velocities, $v_{\theta}/v_0 \in [0, 200]$, are considered for the wettability pattern, Eq.~(\ref{eq:sinetheta}). 
A few comments on these ranges are noteworthy at this point. 
The wavelengths are obviously limited from above by the size of the computational domain. On the other extreme, too short wavelengths are prevented by the energetic cost of accomodating droplets on too small patches, since the local contact angle (that increases with decreasing $\lambda$~\cite{SuppMat}) would possibly exceed the largest substrate contact angle, such that the patterning itself would loose meaning. 
It has been, in fact, proven, both theoretically and experimentally, that for the confinement to be effective, the pattern characteristic size should not be much smaller than the spinodal length scale on the less wettable substrate~\cite{karguptaMorphologicalSelforganizationDewetting2002,karguptaInstabilityPatternFormation2000,nisatoExcitationSurfaceDeformation1999,karimPhaseSeparationUltrathin1998}.

@@

The pattern speeds, on the other hand, are limited by the intrinsic LB bound of the lattice speed of sound~\cite{succiLatticeBoltzmannEquation2001}.\\
The Reynolds number, $Re$, is of the order of $Re \sim 10^{-2}$ on the static substrate and never exceeds the value $\approx 0.2$ in the time dependent case.

@@

\section{Results} 
We start our analysis studying, first, how the rupture times depend on the parameters characterising the wettability pattern, namely the wavelength of the contact angle variation, $\lambda$, and wave speed, $v_{\theta}$.

@@

It is conveyed that, overall, rupture occurs earlier on the static substrate, suggesting that the time variation tends to stabilize the film, in agreement with linear stability analysis results~\cite{sumanDynamicsThinLiquid2006}.

@@

Let us first notice that the film rupture consists of two processes, each with its associated time scale: the growth of unstable interface perturbation with rate $t_{\theta}^{-1}$, and the retraction of liquid from more hydrophobic regions, occurring in a characteristic time $t_R$. 
For longer pattern wavelengths, the dewetting instability amplifies more slowly, such that it is reasonable to take it as the process determining the rupture time, $\tau_r \sim t_{\theta}$.
In this case, from the linearized thin-film equation (in one spatial dimension, for simplicity), obtained setting $h=h_0 + \delta h$ with $\delta h \ll h_0$, we can easily see that the exponential growth of the height perturbation is affected by the wettability pattern (variable contact angle) in such a way that $\partial_t (\delta h) \propto (\partial_x^2 (\partial_h\Pi(h_0))) \delta h$. 
Therefore, since the characteristic time $t_{\theta}$ can be estimated dimensionally as $t_{\theta} \sim \delta h/\dot{(\delta h)}$, the rupture time should go as

@@

The latter is related to the time the liquid takes to flow out of regions of high contact angle, whose size is $\sim \lambda$. 
Hence we have 

@@

    \caption{Rupture times, $\tau_r$, as a function of the pattern wavelength $\lambda$, for $v_{\theta}=0$ (\textcolor{jlblue}{$\bullet$}) and $v_{\theta}=20 v_0$ (\textcolor{jlorange}{$\star$}).
    The continuous and dashed lines indicate the linear, $\sim \lambda$, and quadratic, $\sim \lambda^2$, scaling laws, respectively.}

@@

    \caption{MAIN PANEL: Time evolution of the height fluctuations, $\Delta h(t)$, during the dewetting process on the patterned substrate given by Eq.~(\ref{eq:sinetheta}) with $v_{\theta}= 0$ and $\lambda= 512 h_0$ (blue circles), $\lambda=256 h_0$ (orange squares) and $\lambda=170 h_0$ (green stars).

@@

    INSET: Number of droplets, $N(t)$, as a function of time. 
    The three horizontal dashed lines indicate the number of minima of Eq.~(\ref{eq:sinetheta}), which is $2\left(\frac{L}{\lambda}\right)^2$. 
    In the insets, we display snapshots of the stationary droplet states as grey-scale images of the film thickness field, $h(\mathbf{x},t)$.}

@@

In the main panel of the same figure, we see that the height fluctuations $\Delta h(t) = \max_{\mathbf{x}}\{h(\mathbf{x},t)\}-\min_{\mathbf{x}}\{h(\mathbf{x},t)\}$ grow in time until film rupture and then settle to a constant value, which represents a measure of the mean droplet height $h_d$ (since
droplets are essentially monodisperse), decreasing with the pattern wavelength (as expectable, due to the decrease of the droplet volume, $V_d = \frac{h_0 \lambda^2}{2}$).\\

@@

For $v_{\theta} = 0.2 v_0$ we still observe the formation of droplets, similarly to the stationary case ($v_{\theta} = 0$). 
However, these are transported with the contact angle minima, reproducing a somehow similar behaviour recently described in a numerical study of a droplet on a moving wettability step~\cite{grawitterSteeringDropletsSubstrates2021}.

@@

In order to better characterize the various morphologies we apply the theory of Minkowski's functionals. 
In particular, we employ the second order Minkowski structure metric, $q_2$~\cite{mickelShortcomingsBondOrientational2013, schallerPapaya22DIrreducible2020}, which can be computed from a Voronoi tessellation of the set of discrete points $(x_i, y_i)$ on the 2D lattice, such that the height field lies above a certain threshold\footnote{The expression is: $q_2 = \frac{1}{N}\sum_j \frac{1}{P_j}|\sum_k L^{(j)}_k e^{2i\phi^{(j)}_k}|$, where the inner sum runs over the edges of length $L^{(j)}_k$, of the $j$-th Voronoi cell, whose perimeter is $P_j$, and $\phi^{(j)}_k$ is the polar angle of the normal to the $k$-th edge. 
The outer sum represents an ensemble average over the $N$ points in the set.}. 
The $q_2$ metric quantifies the degree of anisotropy of the dewetting morphology, so it takes relatively large values if the structures formed display a preferential direction. 
Measuring $q_2$ then enables us to clearly distinguish between the formation of droplets and rivulets: much larger $q_2$ values are attained for the latter type of structure, as we can see in Fig.~\ref{fig:msm_q2}.
We observe, on the other hand, that such rivulets are metastable and eventually break up into droplets, as indicated by the collapse of $q_2$ at later times. 
Notice, though, that the $q_2$ signal for any $v_{\theta} >0$ always stays above the one for the static case, suggesting that even the smallest pattern velocity introduces a sizeable deformation of the spherical cap shape.
\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Figure_5.pdf}
    \caption{MAIN PANEL: Rivulets life-times, $\tau_{\text{riv}}$, for various $\Gamma$ values.
    The dashed line is plotted as a guide to the eye to highlight the logarithmic dependence, in agreement with the theoretical prediction, Eq.~(\ref{eq:rivlt}).
    INSET: Height fluctuations, $\Delta h(t)$, vs time, computed along the rivulet axis, for three different $\Gamma$ values.
    }
    \label{fig:stab_ligs_lam2}
\end{figure}
The breakup is the result of a varicose mode of the rivulet~\cite{diezBreakupFluidRivulets2009, mechkovStabilityLiquidRidges2008}, whose wavelength is $\approx\lambda$, such that only $N_{\infty}/2$ droplets are counted after breakup. 
These droplets show a peculiar dynamics, characterized by a periodic sequence of spreading and retraction, driven by the pattern, that we dub "pumping state"~\footnote{see movie file ligament\_formation\_and\_breakup.mp4 in~\cite{SuppMat}.}.

@@

If $U_{\Theta}$ is large as compared to $v_{\theta}$, the film retraction is faster than the transport due to the contact angle field and thus droplets from.
However, if $v_{\theta}$ is larger than $U_{\Theta}$, then the retracting film has too little time to form droplets and ends up in the metastable rivulet state. 
It appears, therefore, natural to consider the ratio of these two velocities, 

@@

The rivulet itself is, in fact, prone to dewetting, with the liquid accumulating over patches around contact angle minima. 
However, as the pattern moves, the instability is tamed due to configurations whereby higher contact angle regions underlie height field maxima, thus tending to revert the fluid flow.
Heuristically speaking, this means that, if we evaluate $\Delta h(t)$ restricted on the rivulet axis, this should grow exponentially (with a certain growth rate $\alpha \propto t_0$) only when the system is in the unstable configuration. 
